An Open and Continuously Updated Fern Tree of Life (FTOL)

### 2022-04-01 ms/manuscript.Rmd
Update MS

---
title: "An Open and Continuously Updated Fern Tree of Life (FTOL)"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: yes
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "frontiers.csl"]
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template_SI.docx
header-includes: 
  - \usepackage{float} # keep figures from floating away
  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # keep figures from floating away
  - \usepackage{fancyhdr} # enable header
  - \cfoot{\thepage} # footer
editor_options: 
  chunk_output_type: console
params:
  doc_type: manual
---

@@

# Specify path to FTOL cache
# FIXME: change to ftol_cache when ready
ftol_cache <- here::here("working/_targets")

tar_load(
  c(
    accs_exclude,
    con_fossil_calibration_tips,
    fern_sanger_seqs_raw,
    plastid_calibration_dates,
    plastome_alignment,
    plastome_alignment_tbl,
    plastome_metadata_raw_all,
    plastome_metadata_renamed,
    plastome_metadata_renamed,
    plastome_seqs_combined_filtered,
    plastome_tree_rooted,
    ppgi_taxonomy,
    pteridocat,
    raw_fasta_all,
    sanger_accessions_selection,
    sanger_alignment,
    sanger_alignment_tbl,
    sanger_con_tree_dated,
    sanger_con_tree_rooted,
    sanger_seqs_combined_filtered,
    sanger_seqs_rogues_removed,
    ts_fossil_calibration_tips
  ),
  store = ftol_cache
)

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    bb_coverage_by_rank,
    bb_total_coverage,
    coverage_by_rank,
    crown_ages,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    fossil_ferns_raw,
    ftol_ts_comp_mod,
    gb_species_by_year,
    other_dates,
    plastome_iqtree_log,
    ref_files,
    sanger_iqtree_log_2000,
    sanger_sampling,
    total_coverage,
    treepl_cv_results
  ),
  store = here::here("_targets")
)
```

@@

Ferns, with about `r number(12000)` species, are the second most diverse clade of vascular plants after seed plants.
Ferns have received intense focus from molecular systematists, resulting in the publication of molecular phylogenies for every major clade and DNA sequences for more than one third of all fern species.
Global fern phylogenies have been published periodically, but as molecular systematics research continues at a rapid pace, these become quickly outdated.

Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from DNA sequence data available via GenBank.
Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with commonly sequenced plastid regions (many taxa, few loci) to obtain a global, species-level fern phylogeny with high resolution along the backbone and maximal sampling across the tips.
We use an expert-curated reference taxonomy to resolve synonyms in general compliance with Pteridophyte Phylogeny Group I (PPG I), the community-driven taxonomic system widely accepted by many pteridologists.

The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of `r ca` 40% relative to the previous largest fern phylogeny.
Using an updated and expanded list of `r n_calibrations_ferns` fern fossil calibration points, we find estimated ages for most families and deeper clades to be significantly older than earlier studies.

FTOL and its accompanying datasets including the fossil list and taxonomic database will be updated on a regular basis and made available via a web portal and R packages, enabling immediate access to the most up-to-date, comprehensively-sampled fern phylogeny.
FTOL will be useful for anybody studying this important group of plants over a wide range of taxonomic scales, from smaller clades to the entire tree.
We anticipate FTOL will be particularly relevant for macroecological studies at regional to global scales and will inform future taxonomic systems with the most recent hypothesis of fern phylogeny.

@@

Ferns (ca. `r number(12000)` species) are the second most diverse clade of vascular plants after seed plants (ca. `r number(300000)` species) and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
Key to any investigation of evolutionary biology in this group is a well-sampled phylogeny.
Fortunately, ferns have received relatively intense focus from molecular systematists, which has resulted in the publication of trees for all major clades and DNA sequence data for more than one third of all currently recognized fern species. 
Thus, there is both sufficient sampling and a pressing need for a globally sampled fern phylogeny.

Past efforts to construct such a global phylogeny have steadily expanded their sampling as sequences have been added to GenBank [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
Indeed, growth of of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny of ferns [@Testo2016a\; `r figure("genbank")`].
There is a need therefore, not only for a global fern phylogeny, but one that is continuously updated to keep pace with the rapid accumulation of molecular data.
This would eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.

Multiple frameworks have been put forth to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
While these may be appropriate in some cases, they are not expected to produce an optimal fern phylogeny due to the need to take a "one-size-fits-all" approach to accommodate such a wide phylogenetic breadth.
By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can then be used "as-is" by biologists studying these organisms.
Furthermore, there is much to be gained from integrating a carefully designed global fern phylogeny with the fern systematics community that would not be as easily accomplished with a "tree of all life" or "tree of all plants".

Recently, a genus-level taxonomy of ferns and lycophytes was established using an inclusive, community-driven approach  [@PteridophytePhylogenyGroupI2016\; hereafter "PPG I"].
PPG I has been widely accepted and used, but there were problematic (non-monophyletic) genera included at the time of publication, and many taxonomic changes have been (and will continue to be) proposed since [`r eg`, @Almeida2017; @Shang2018; @Zhang2020].
The next iteration of PPG (`r ie`, PPG II) will ideally be an online, open resource that can be updated as necessary.
We envision an open, continuously updated global fern phylogeny that could be directly integrated with PPG II such that taxonomic decisions can be made based on community consensus and the most recently available data.

Here, we leverage specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL; `r figure("workflow")`).
We plan to run the pipeline on a regular basis and make the results freely available online through a web portal and R package.
This will enable anybody interested in the biology of ferns to have access to the most recent hypothesis of fern phylogeny.
We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern phylogeny; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; and 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.

@@

The plastid genome has been the most widely sequenced genomic compartment in ferns by far (`r figure("genbank")`), so we used plastid loci to build our tree.
Our sampling includes two major categories of sequence data: 1) seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics of ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci"); and 2) a much larger set of single-copy loci typically obtained from next-generation sequencing of the plastome ("plastome loci").
Here, the *trnL*--*trnF* locus includes the *trnL* intron, the 3' *trnL* exon, and the *trnL*--*trnF* intergenic spacer; the *rps4*--*trnS* locus includes only the *rps4*--*trnS* intergenic spacer.
The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes, including *atpA*, *atpB*, *matK*, *rbcL*, and *rps4*), and then combined with the *trnL*--*trnF* and *rps4*--*trnS* loci (79 loci total).
Here, "locus" refers to an individual gene, intergenic spacer, or a unit comprised of these in the case of *trnL*--*trnF*.

@@

All sequences were downloaded from GenBank.
GenBank queries were formatted to include locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
There is no formal definition of genomic vs. Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) or plastome sequences (> `r number(7000)` bp) with the "[SLEN]" term.

@@

This project aimed to generate a phylogeny that is consistent with PPG I, while accounting for taxonomic changes that have been made since the publication of PPG I.
Species names in GenBank, which use the NCBI taxonomy [@Federhen2012; @Schoch2020], do not necessarily conform to PPG I.
Furthermore, the NCBI taxonomy is not curated specifically for ferns and includes many fern synonyms.
Therefore, we standardized all species names in the GenBank sequences against a newly generated reference taxonomy.
Our reference taxonomy is based on the World Ferns database v.12.8 [@Hassler2022], which conforms to PPG I (with some exceptions explained below) and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].

@@

We manually inspected any fuzzy matched or non-matching names.
This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
We thus updated and edited the initial World Ferns database using the dwctaxon R package [@joel_nitta_2022_6388272], which is designed to work with taxonomic data in the Darwin Core format.
We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL [@joel_nitta_2022_6388787].

@@

Another potential issue with GenBank sequences is the presence of misidentified sequences, poor quality sequences, and contaminations (hereafter collectively referred to as "rogues").
We removed putative rogues from Sanger sequences as follows.
First, we constructed a BLAST library including all extracted Sanger sequences.
Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
We compared the families (following PPG I) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, that query was considered a contaminant and excluded from further analysis.
For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family.
Species belonging to monotypic families were not considered for this filtering.
While this method cannot account for rogues at finer taxonomic levels, it is an efficient approach for removing obviously erroneous sequences.
We inspected all accessions flagged this way as rogues before excluding them.
In cases where the family mismatch was due to incorrect taxonomy (`r eg`, the species name in GenBank matches the correct family but the name used in the taxonomic database does not), we updated the taxonomic database accordingly.

@@

A typical step in multilocus phylogenetic workflows is to concatenate loci across samples.
For phylogenetic studies aiming to include one tip per species, this is often done by concatenating the longest sequence per locus within each species.
However, such an approach is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions may have multiple loci per accession, but we already split those out using superCRUNCH as described previously).
First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
Then, we concatenated loci that met any of the following three conditions: 1) if a species was monophyletic, loci were concatenated by selecting the longest accession per locus within that species; 2) if all accessions originated from the same voucher specimen, loci were concatenated across accessions; 3) if all accessions for a given species originated from only one publication, loci were concatenated across accessions.
Accessions not meeting any of these conditions were not concatenated.
All calculations of sequence length excluded missing bases ("?", "N", or "-").

We selected the final set of concatenated loci for each species based on presence of *rbcL* and sequence length.
We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and maximal sampling of one locus has been shown to improve results in super-matrix phylogenies [@Talavera2021].
Our procedure for selecting accessions for each species is as follows.

1. If accessions are concatenated including *rbcL* and at least one other locus, select the set of concatenated accessions with the greatest total sequence length.
2. Otherwise, if accessions include *rbcL* only, select the accession with the longest *rbcL* sequence.
3. Otherwise, select the accession or set of concatenated accessions with the greatest total sequence length.

The above steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
For these, we selected one specimen per species with the longest combined sequence length across all plastome loci.

@@

For non-spacer regions (genes), we aligned each locus separately in MAFFT with automatic adjustment for sequence direction and other settings on default. Spacer regions are difficult to align across higher taxonomic levels within ferns (`r eg`, family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer scales (`r eg`, within genera or family) where slower-evolving genes like *rbcL* may not provide enough resolution.
Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "\-\-merge" option.
This retains the indels within each subalignment while aligning across subalignments.
For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "\-\-merge".
Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data (*trnL*--*trnF* and *rps4*--*trnS*).
We also excluded outgroups from the spacer region data as they cannot be reliably aligned to ferns.

@@

We inspected the initial Sanger phylogeny to identify any remaining rogues or names that needed updating in the taxonomic database.
We checked for monophyly at taxonomic levels at or above genus using the R package "MonoPhy" [@Schwery2016].
We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect [`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus\; expected monophyly follows @PteridophytePhylogenyGroupI2016].
However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision and are known to include non-monophyletic genera (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants [@Turland2018].

Based on the results of this inspection, we updated the list of rogue accessions to be excluded, updated the taxonomic database, and then re-ran all analyses up to this step.
During each iteration of the analysis, it is possible that the GenBank accessions added in place of the excluded rogues themselves include rogues.
Therefore, we repeated this process until the monophyly of all higher-level (`r eg`, genus rank and higher) taxa that were expected to be monophyletic according to @PteridophytePhylogenyGroupI2016 was either confirmed or could not be achieved due to outstanding taxonomic issues.

@@

We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset, with the same model selection procedure and bootstrapping as used for the backbone phylogeny. IQ-TREE was initially run using 1,000 iterations (default), but at the end of this run the bootstrap correlation coefficient of split occurrence frequencies was 0.96, which is below the threshold for convergence (0.99). We then continued the search for another 1,000 iterations (2,000 total), but the correlation coefficient fluctuated between 0.96 and 0.98 without overall improvement and again failed to converge. This suggests that the search was stuck in local optima. As our alignment contains many species with very similar sequences, it is likely that the search algorithm is unable to optimize many highly similar topologies that only vary in positions of closely related species at the tips. Considering that additional iterations were unlikely to converge, we therefore conducted 10 independent runs of IQTREE with 1,000 iterations each, and selected the run with the best (highest) combined log-likelihood of the ML and consensus trees [@Zhou2018].

@@

Dating was conducted separately on the ML tree and the consensus tree.
We rooted the tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].

>>>>

---
title: "An Open and Continuously Updated Fern Tree of Life (FTOL)"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: yes
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "frontiers.csl"]
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template.docx
header-includes: 
  - \usepackage{float} # keep figures from floating away
  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # keep figures from floating away
  - \usepackage{fancyhdr} # enable header
  - \cfoot{\thepage} # footer
editor_options: 
  chunk_output_type: console
params:
  doc_type: manual
---

@@

# Specify path to FTOL cache
ftol_cache <- here::here("ftol_cache")

tar_load(
  c(
    accs_exclude,
    con_fossil_calibration_tips,
    fern_sanger_seqs_raw,
    fossil_ferns_raw,
    plastid_calibration_dates,
    plastome_alignment,
    plastome_alignment_tbl,
    plastome_metadata_raw_all,
    plastome_metadata_renamed,
    plastome_metadata_renamed,
    plastome_seqs_combined_filtered,
    plastome_tree_rooted,
    ppgi_taxonomy,
    pteridocat,
    raw_fasta_all,
    sanger_accessions_selection,
    sanger_alignment,
    sanger_alignment_tbl,
    sanger_con_tree_dated,
    sanger_con_tree_rooted,
    sanger_seqs_combined_filtered,
    sanger_seqs_rogues_removed,
    ts_fossil_calibration_tips
  ),
  store = ftol_cache
)

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    bb_coverage_by_rank,
    bb_total_coverage,
    coverage_by_rank,
    crown_ages,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    ftol_ts_comp_mod,
    gb_species_by_year,
    other_dates,
    plastome_iqtree_log,
    ref_files,
    sanger_iqtree_log_2000,
    sanger_sampling,
    total_coverage,
    treepl_cv_results
  ),
  store = here::here("_targets")
)
```

@@

Ferns, with about `r number(12000)` species, are the second most diverse clade of vascular plants after angiosperms.
They have been the subject of numerous molecular phylogenetic studies, resulting in the publication of trees for every major clade and DNA sequences from nearly half of all species.
Global fern phylogenies have been published periodically, but as molecular systematics research continues at a rapid pace, these become quickly outdated.

Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from DNA sequence data available in GenBank.
Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with commonly sequenced plastid regions (many taxa, few loci) to obtain a global, species-level fern phylogeny with high resolution along the backbone and maximal sampling across the tips.
We use an curated reference taxonomy to resolve synonyms in general compliance with the community-driven Pteridophyte Phylogeny Group I classification.

The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of `r ca` 40% relative to the most recently published global fern phylogeny.
Using an updated and expanded list of `r n_calibrations_ferns` fern fossil constraints, we find estimated ages for most families and deeper clades to be considerably older than earlier studies.

FTOL and its accompanying datasets, including the fossil list and taxonomic database, will be updated on a regular basis and are available via a web portal (https://fernphy.github.io) and R packages, enabling immediate access to the most up-to-date, comprehensively sampled fern phylogeny.
FTOL will be useful for anybody studying this important group of plants over a wide range of taxonomic scales, from smaller clades to the entire tree.
We anticipate FTOL will be particularly relevant for macroecological studies at regional to global scales and will inform future taxonomic systems with the most recent hypothesis of fern phylogeny.

@@

Ferns (ca. `r number(12000)` species) are the second most diverse clade of vascular plants after angiosperms (ca. `r number(300000)` species) and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
Key to any investigation of evolutionary history in this group is a well-sampled phylogeny.
Fortunately, ferns have received relatively intense focus from molecular systematists, which has resulted in the publication of trees for all major clades and DNA sequence data from nearly half of all currently recognized fern species.
Thus, there is both a pressing need and sufficient sampling for a globally sampled fern phylogeny.

Past efforts to construct such a global phylogeny have steadily expanded their sampling, at first by mostly generating new sequences, then later by mining GenBank [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
Indeed, growth of of plastid fern accessions in GenBank show no sign of slowing since the most recent global fern phylogeny [@Testo2016a\; `r figure("genbank")`].
There is a need therefore, not only for a revised global fern phylogeny, but also one that is continuously updated to keep pace with the rapid accumulation of molecular data going forward.
Such an effort would eliminate the need for researchers to "rebuild the wheel" each time the need for a globally sampled fern phylogeny arises.

Multiple frameworks have been put forth to automatically or semi-automatically generate trees for any particular part of the tree of life [@Antonelli2016], all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015], which would of course subsume a global fern phylogeny.
While such approaches are well-suited to some studies, they cannot be expected to produce an optimal fern phylogeny due to the use of "one-size-fits-all" methods to accommodate such a wide phylogenetic breadth.
By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can then be used "as-is" by biologists studying these organisms.
Furthermore, there is much to be gained from integrating a carefully designed global fern phylogeny with the fern systematics community that would not be as easily accomplished with a "tree of all life" or "tree of all plants".

Recently, a genus-level taxonomy of ferns and lycophytes was established using an inclusive, community-driven approach  [@PteridophytePhylogenyGroupI2016\; hereafter "PPG I"].
PPG I has been widely accepted and used, but there were problematic (non-monophyletic) genera included at the time of publication, and many taxonomic changes have been (and will continue to be) proposed since [`r eg`, @Almeida2017; @Shang2018; @Zhang2020].
The next iteration of the PPG classification (`r ie`, PPG II) will ideally be an online, open resource that can be updated as necessary.
We envision an open, continuously updated global fern phylogeny that could be directly integrated with PPG II such that taxonomic decisions can be made based on community consensus and the most recently available data.

Here, we leverage taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL; `r figure("workflow")`).
We plan to run the pipeline on a regular basis and make the results freely available online through a web portal (https://fernphy.github.io) and R package [@ftol2022c].
This will enable anybody interested in the biology of ferns to have access to the most current hypothesis of fern phylogeny, as well as an associated time-tree dated using a curated list of fern fossils.
We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern relationships; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; and 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.

@@

The plastid genome has been the most widely sequenced genomic compartment in ferns by far (`r figure("genbank")`), so we used plastid loci to build our tree.
Our sampling includes two major categories of sequence data: 1) seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used in molecular analyses of ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci"); and 2) a much larger set of single-copy loci typically obtained from next-generation sequencing of the plastome ("plastome loci").
Here, the *trnL*--*trnF* locus includes the *trnL* intron, the 3' *trnL* exon, and the *trnL*--*trnF* intergenic spacer; the *rps4*--*trnS* locus includes only the *rps4*--*trnS* intergenic spacer.
The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes, including *atpA*, *atpB*, *matK*, *rbcL*, and *rps4*), and then combined with the *trnL*--*trnF* and *rps4*--*trnS* loci (79 loci total).
Here, "locus" refers to an individual gene, intergenic spacer, or a unit comprised of these in the case of *trnL*--*trnF*.

@@

All sequences were downloaded from GenBank.
GenBank queries were formatted to include the locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify a date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query results in the same set of sequences regardless of when the query is made (with the possible exception of accessions that were embargoed at the time of querying).
Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
There is no formal definition of genomic vs. Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) and plastome (> `r number(7000)` bp) accessions with the "[SLEN]" term.

@@

This project aimed to generate a phylogeny that is consistent with PPG I, while accounting for taxonomic changes that have been made since its publication.
Species names in GenBank, which use the NCBI taxonomy [@Federhen2012; @Schoch2020], do not necessarily conform to PPG I.
Furthermore, the NCBI taxonomy is not curated specifically for ferns and includes many fern synonyms.
Therefore, we standardized all species names in the GenBank sequences against a newly generated reference taxonomy.
Our reference taxonomy is based on the World Ferns database v.12.8 [@Hassler2022], which conforms to PPG I (with some exceptions explained below) and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].

@@

We manually inspected any fuzzily matched or non-matching names.
This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
We thus updated and edited the initial World Ferns database using the dwctaxon R package [@joel_nitta_2022_6388272], which is designed to work with taxonomic data in the Darwin Core format.
We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL [@ftol2022a].

@@

Another potential issue with GenBank sequences is the presence of misidentified sequences, poor quality sequences, and contaminants (hereafter collectively referred to as "rogues").
We removed putative rogues from Sanger sequences as follows.
First, we constructed a BLAST library including all extracted Sanger sequences.
Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
We compared the families (following PPG I) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, , that query was considered a rogue and excluded from further analysis.
For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family to avoid false positives.
Species belonging to monotypic families were not considered for this filtering.
While this method cannot account for rogues at finer taxonomic levels, it is an efficient approach for removing obviously erroneous sequences.
We inspected all accessions flagged this way as rogues before excluding them.
In cases where the family mismatch was due to incorrect taxonomy (`r eg`, the species name in GenBank matches the correct family but the name used in the taxonomic database does not), we updated the taxonomic database accordingly.

@@

A typical step in multilocus phylogenetic workflows is to concatenate loci across samples.
For phylogenetic studies aiming to include one tip per species, this is often done by concatenating the longest sequence per locus within each species, regardless of source.
However, such an approach is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions may contain multiple loci, but we already split those out using superCRUNCH as described above).
First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
Then, we concatenated loci that met any of the following three conditions: 1) if a species was monophyletic, loci were concatenated by selecting the longest accession per locus within that species; 2) if all accessions originated from the same voucher specimen, loci were concatenated across accessions; 3) if all accessions for a given species originated from only one publication, loci were concatenated across accessions.
Accessions not meeting any of these conditions were not concatenated.
All calculations of sequence length excluded missing bases ("?", "N", or "-").

We selected the final set of concatenated loci for each species based on presence of *rbcL* and sequence length.
We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and maximal sampling of one locus has been shown to improve results in super-matrix phylogenies [@Talavera2021].
Our procedure for selecting accessions for each species is as follows: 1) if accessions are concatenated including *rbcL* and at least one other locus, select the set of concatenated accessions with the greatest total sequence length; 2) otherwise, if accessions include *rbcL* only, select the accession with the longest *rbcL* sequence; 3) otherwise, select the accession or set of concatenated accessions with the greatest total sequence length.
These steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
For these, we selected the one specimen per species with the longest combined sequence length across all plastome loci.

@@

For non-spacer regions (genes), we aligned each locus separately in MAFFT with automatic adjustment for sequence direction and other settings on default. Spacer regions are difficult to align across higher taxonomic levels within ferns (`r eg`, family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer scales (`r eg`, within genera or family) where slower-evolving genes like *rbcL* may not provide enough resolution.
Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "\-\-merge" option.
This retains the indels within each subalignment while aligning across subalignments.
Sequences from families with fewer than three species each could not be used for subalignments, so these were added as "singletons" during the MAFFT "\-\-merge".
Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data (*trnL*--*trnF* and *rps4*--*trnS*).
We also excluded outgroups from the spacer region data as they cannot be reliably aligned to ferns.

@@

We inspected the initial Sanger phylogeny to identify any remaining rogues or names that needed updating in the taxonomic database.
We checked for monophyly at taxonomic levels at or above genus using the R package "MonoPhy" [@Schwery2016].
We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect (`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus; expected monophyly follows PPG I, 2016).
However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision and are known to include non-monophyletic genera (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that were not already validly published according to the International Code of Nomenclature for Algae, Fungi, and Plants [@Turland2018].
Some unpublished names appearing in GenBank and World Ferns databases were not excluded.

Based on the results of this inspection, we updated the list of rogue accessions to be excluded, updated the taxonomic database, and then re-ran all analyses up to this step.
During each iteration of the analysis, it is possible that the GenBank accessions added in place of the excluded rogues themselves include rogues.
Therefore, we repeated this process until the monophyly of all higher-level (`r eg`, genus rank and higher) taxa that were expected to be monophyletic according to PPG I (2016) was either confirmed or could not be achieved due to outstanding taxonomic issues.

@@

We generated the final Sanger phylogeny using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset, with the same model selection procedure and bootstrapping as used for the backbone phylogeny.
IQ-TREE was initially run using 1,000 iterations (default), but at the end of this run the bootstrap correlation coefficient of split occurrence frequencies was 0.96, which is below the threshold for convergence (0.99).
We then continued the search for another 1,000 iterations (2,000 total), but the correlation coefficient fluctuated between 0.96 and 0.98 without overall improvement or convergence.
This suggests that the search was stuck in a local optimum.
As our alignment contains many species with very similar sequences, it is likely that the search algorithm is unable to optimize many highly similar topologies that only vary in positions of closely related species at the tips.
Considering that additional iterations were unlikely to converge, we therefore conducted 10 independent runs of IQTREE with 1,000 iterations each, and selected the run with the best (highest) combined log-likelihood of the ML and consensus trees [@Zhou2018].

@@

Dating was conducted separately on the ML tree and the consensus tree.
We rooted the tree with bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].

We selected `r n_calibrations_ferns` fern fossils to use as constraints from a newly curated database of `r n_fern_fossils_total` fern fossil taxa called "ferncal" [@ftol2022b\; `r s_table("fossils")`, Appendix S2].
The full ferncal database is available at https://fernphy.github.com/ferncal.
In the case of redundant fossils (those assignable to the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil). 
We assigned fossils to lineages only after consulting the original publication (as opposed to simply relying on the fossil name) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
Taxonomic concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be constrained with a given fossil could change.
We applied one fossil constraint outside of ferns (stem euphyllophytes; `r fossils$Euphyllophytes$age` Ma), and fixed the root age of the tree (land plants) at 475 Ma as in @Testo2016a and @Qi2018.
All constraints other than the root are minimum ages (`r s_figure("fossil_map")`).
